---
title: "RepDataPeer1"
author: "Esteban R. Freyre"
date: "8/5/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```
Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement -- a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Exploratory Analysis

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(DT)
```
Code for reading in the dataset and/or processing the data
download data set from "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

```{r}
activity<-read.table("activity.csv",sep = ",",header = TRUE,colClasses = c("numeric","POSIXct","numeric"))
str(activity)
summary(activity)
```
Percentage of NAs in the sample
```{r}
mean(is.na(activity$steps))
```


```{r}
library(DT)
datatable(activity)
```
```{r fig.height=3.5, fig.width=5.5}
library(ggplot2)
qplot(date,steps, data=activity)
```
```{r fig.height=3.5, fig.width=5.5, message=FALSE, warning=FALSE}
library(lattice)
histogram(activity$steps,activity$date)
```
Calculate the total number of steps taken per day
```{r}
daysteps<-mutate(activity,day=as.Date(date))
daysteps<-group_by(daysteps,day)
```
```{r}
stptotals<-summarise(daysteps,stpmean=mean(steps,na.rm = TRUE),
          stpdaily=sum(steps,na.rm = TRUE))
summary(stptotals)
```

Histogram of the total number of steps taken each day
```{r fig.height=3.5, fig.width=5.5}
hist(stptotals$stpdaily, col="green", breaks = 60, main = "Total Steps per Day",xlab = " Daily Steps")
```

Average daily Activity Pattern  
```{r}
inter<-activity%>%group_by(interval)%>%summarise(meansteps=mean(steps,na.rm = TRUE),
                                                 totalsteps=sum(steps,na.rm = TRUE),
                                                 maxsteps=max(steps,na.rm = TRUE))
```
Time series plot of the 5 min average number of steps taken  
```{r fig.height=3.5, fig.width=5.5}
with(inter,plot(interval,meansteps,type="l",col="blue",main = "Mean Steps per Interval",xlab = "Five Minutes Interval",ylab = "Average Steps"))
abline(v=835, col="red")
```
Mean and median number of steps taken each day  
```{r}
daystats<-summarise(stptotals,stpmean=mean(stpdaily,na.rm = TRUE),
          stpmedian=median(stpdaily,na.rm = TRUE))
daystats
```
The 5-minute interval that, on average, contains the maximum number of steps is interval 835. See graphic above for intervals.
```{r}
maxinter<-arrange(inter,desc(meansteps))
head(maxinter,10)
```

## Imputing missing values  

Calculate and report the total number of missing values in the datase  
```{r}
sum(is.na(activity))
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
The function "replace_na" from tidyr package will be used.

Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” 
indicating whether a given date is a weekday or weekend day.
```{r}
weekday1<-activity$date
str(weekday1)
summary(weekday1)
```
```{r}
weekday2<-weekdays.POSIXt(weekday1)
str(weekday2)
tail(weekday2)
```
```{r}
weekday2<-gsub("Saturday|Sunday","weekend",weekday2)
weekday2<-gsub("Monday|Tuesday|Wednesday|Thursday|Friday","weekday",weekday2)
str(weekday2)
weekday3<-as.factor(weekday2)
str(weekday3)
```
```{r}
activity$weekday<-weekday3
str(activity)
summary(activity)
sum(is.na(activity))
```

Filling NAs values by using the weekday average.
```{r}
activitynona<-activity
summary(activitynona)
```

```{r}
activitynona[is.na(activitynona)]<-35
summary(activitynona)
str(activitynona)
sum(is.na(activitynona))
```

```{r}
group_by(activitynona,weekday)%>%summarise(weekdaymean=mean(steps,na.rm =TRUE),
                                   weeekdaytotal=sum(steps,na.rm =TRUE))
```   
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Changes in average daily steps per day comparison  
```{r}
dailytotalsnona<-group_by(activitynona,date)%>%summarise(dailytotals=sum(steps,na.rm=TRUE),
dailymean=mean(steps,na.rm = TRUE))
```
Histogram of the total number of steps taken each day
```{r fig.height=6}
par(mfcol=c(2,1))
hist(stptotals$stpdaily, col="green", breaks = 60, main = "Total Steps per Day w/ NAs",xlab = " Daily Steps")
hist(dailytotalsnona$dailytotals, col="lightgreen", breaks = 60, main = "Total Steps per Day w/out NAs",xlab = " Daily Steps")
```
The main difference is that the null steps frequency is lower.

```{r fig.height=6}
par(mfcol=c(2,1))
with(stptotals,plot(day,stpmean,type="l",col="magenta",main = "Mean Steps per Day",xlab = "Date",ylab = "Average Daily Steps"))
with(dailytotalsnona,plot(date,dailymean,type="l",col="red",main = "Mean Steps per Day- NAs Replaced",xlab = "Date",ylab = "Average Daily Steps"))
```

                                                              
Make a panel plot containing a time series plot (i.e. type = "l"\color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r}
f<-factor(weekday,labels = c("weekday","weekend"))

xyplot(dailytotal ~ day | f,week,type="l",layout = c(1, 2))
```

```{r}
days<-mutate(activitynona,day=date)
totalsc<-group_by(days,day)%>%summarise(dailymean=mean(steps,na.rm = FALSE),   dailymedian=median(steps,na.rm = TRUE),
          dailytotal=sum(steps,na.rm = TRUE),
          dailyinterval=sum(interval,na.rm = TRUE))
summary(totalsc)
str(totalsc)
```
Changes in histograms comparison  
```{r fig.height=6}
par(mfcol=c(2,1))
hist(stptotals$stpdaily, col="green", breaks = 60, main = "Total Steps per Day",xlab = " Daily Steps")
hist(totalsc$dailytotal, col="lightgreen", breaks = 60, main = "Total Steps per Day",xlab = " Daily Steps")
```

Changes in average daily steps per day comparison  
```{r fig.height=6}
par(mfcol=c(2,1))
with(stptotals,plot(day,stpmean,type="l",col="magenta",main = "Mean Steps per Day",xlab = "Date",ylab = "Average Daily Steps"))
with(totalsc,plot(day,dailymean,type="l",col="darkred",main = "Mean Steps per Day w NAs Replaced",xlab = "Date",ylab = "Average Daily Steps"))
```
Make a panel plot containing a time series plot (i.e. type = "l"\color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r}
week<-mutate(totalsc,weekday=weekdays.POSIXt(day))
str(week)
summary(week)
```

```{r}
weekday<-week$weekday
weekday<-gsub("Saturday|Sunday","weekend",weekday)
weekday<-gsub("Monday|Tuesday|Wednesday|Thursday|Friday","weekday",weekday)
str(weekday)
```
```{r}
totalsc$weekday<-as.factor(weekday)
str(totalsc)
sum(is.na(totalsc))
```
```{r}
group_by(week,weekday)%>%summarise(weeekdaymean=mean(dailymean,na.rm =TRUE),
                                     weeekdaytotal=sum(dailytotal,na.rm =TRUE),
                                   weeekendmean=mean(dailymean,na.rm =TRUE),
                                   weekendtotal=sum(dailytotal,na.rm =TRUE))
```              
```{r}
f<-factor(weekday,labels = c("weekday","weekend"))

xyplot(dailytotal ~ day | f,week,type="l",layout = c(1, 2))
```
```{r}
par(mfcol=c(1,1))
activity5<-transform(totalsc,weekday=factor(weekday))

  with(subset(activity5,weekday==weekday(plot(dailytotal~day,type="l",col="red"))
```

```{r}
weekday<-mutate(activity,weekday=weekdays.POSIXt(date))
activity<-weekday¿
str(activity)
```

Submitting the Assignment

To submit the assignment:

    Commit your completed PA1_template.Rmd\color{red}{\verb|PA1_template.Rmd|}PA1_template.Rmd file to the master\color{red}{\verb|master|}master branch of your git repository (you should already be on the master\color{red}{\verb|master|}master branch unless you created new ones)
    Commit your PA1_template.md and PA1_template.html files produced by processing your R markdown file with knit2html() function in R (from the knitr package) by running the function from the console.
    If your document has figures included (it should) then they should have been placed in the figure/ directory by default (unless you overrided the default). Add and commit the figure/ directory to your git repository so that the figures appear in the markdown file when it displays on github.
    Push your master\color{red}{\verb|master|}master branch to GitHub.
    Submit the URL to your GitHub repository for this assignment on the course web site.

In addition to submitting the URL for your GitHub repository, you will need to submit the 40 character SHA-1 hash (as string of numbers from 0-9 and letters from a-f) that identifies the repository commit that contains the version of the files you want to submit. You can do this in GitHub by doing the following

    Going to your GitHub repository web page for this assignment
    Click on the “?? commits” link where ?? is the number of commits you have in the repository. For example, if you made a total of 10 commits to this repository, the link should say “10 commits”.
    You will see a list of commits that you have made to this repository. The most recent commit is at the very top. If this represents the version of the files you want to submit, then just click the “copy to clipboard” button on the right hand side that should appear when you hover over the SHA-1 hash. Paste this SHA-1 hash into the course web site when you submit your assignment. If you don't want to use the most recent commit, then go down and find the commit you want and copy the SHA-1 hash.

    Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays()\color{red}{\verb|weekdays()|}weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

    Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
    Make a panel plot containing a time series plot (i.e. type = "l"\color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

Submitting the Assignment

To submit the assignment:

    Commit your completed PA1_template.Rmd\color{red}{\verb|PA1_template.Rmd|}PA1_template.Rmd file to the master\color{red}{\verb|master|}master branch of your git repository (you should already be on the master\color{red}{\verb|master|}master branch unless you created new ones)
    Commit your PA1_template.md and PA1_template.html files produced by processing your R markdown file with knit2html() function in R (from the knitr package) by running the function from the console.
    If your document has figures included (it should) then they should have been placed in the figure/ directory by default (unless you overrided the default). Add and commit the figure/ directory to your git repository so that the figures appear in the markdown file when it displays on github.
    Push your master\color{red}{\verb|master|}master branch to GitHub.
    Submit the URL to your GitHub repository for this assignment on the course web site.

In addition to submitting the URL for your GitHub repository, you will need to submit the 40 character SHA-1 hash (as string of numbers from 0-9 and letters from a-f) that identifies the repository commit that contains the version of the files you want to submit. You can do this in GitHub by doing the following

    Going to your GitHub repository web page for this assignment
    Click on the “?? commits” link where ?? is the number of commits you have in the repository. For example, if you made a total of 10 commits to this repository, the link should say “10 commits”.
    You will see a list of commits that you have made to this repository. The most recent commit is at the very top. If this represents the version of the files you want to submit, then just click the “copy to clipboard” button on the right hand side that should appear when you hover over the SHA-1 hash. Paste this SHA-1 hash into the course web site when you submit your assignment. If you don't want to use the most recent commit, then go down and find the commit you want and copy the SHA-1 hash.

  